MESOSCOPE manuscript

### 2023-09-06 manuscript/manuscript.Rmd
Fixing up intro, adding section headers, creating supplement, ce

---
title: "MESOSCOPE manuscript results/figures"
author: "William Kumler"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE)
library(tidyverse)

@@

### Metabolome changes significantly across the eddy dipole at DCM and 175m

We found that the metabolome differed between the cyclone and anticyclone samples, with multivariate statistics detecting the largest differences in the DCM and 175 meter samples while the surface was more homogenous (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by SLA alone according to the PERMANOVA (p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the deep samples, 12% of the variance was explained by SLA (p = 0.008) according to the PERMANOVA and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to capture largely variation between biological triplicates. At the surface, SLA trends are less evident with a larger p-value (0.087) and a lower R$^2$ (0.09) (Figure 1).

@@

*Supplemental figure: Results of a principal component analysis performed separately on each depth of the eddy transect metabolome. A) Regressions of SLA against various principal components, broken down by depth. % variance explained by each PC is noted in the upper right corner of the plot. Colors have been added corresponding to the SLA value for clarity but are redundant with the x-axis values. Lines of best fit and the associated standard error have been added behind the points in black and translucent grey, respectively. Strong associations exist between PC1 and SLA in the DCM samples and PC2 and SLA in the deep samples. B) Principal component 1 values for the 175 meter samples at each station showing large variance between triplicates rather than any external factor.*

@@

# Significance testing between depths and SLA classes for % PC in metab
# pc_comp_df %>% aov(formula=frac_in_metab~depth) %>% TukeyHSD()
# pc_comp_df %>% aov(formula=frac_in_metab~sla_class) %>% TukeyHSD()

# Significance testing for differences in % in metabs as a function of SLA by depth
# pc_comp_df %>%
#   nest(data=-depth) %>%
#   mutate(aov_out=map(data, ~broom::tidy(aov(.x$frac_in_metab~.x$sla_class)))) %>%
#   mutate(kw_out=map(data, ~broom::tidy(kruskal.test(.x$frac_in_metab~.x$sla_class)))) %>%
#   pivot_longer(c(aov_out, kw_out)) %>%
#   unnest(value) %>%
#   drop_na(p.value) %>%
#   select(depth, name, p.value) %>%
#   group_by(name) %>%
#   mutate(p.value=p.adjust(p.value, method="fdr")) %>%
#   ungroup() %>%
#   arrange(name, depth)
# pc_comp_df %>%
#   ggplot(aes(x=sla_class, y=frac_in_metab)) +
#   geom_boxplot() +
#   facet_wrap(~depth, nrow=1)

# Permutation testing for effect size test on Deep SLA differences
# set.seed(123)
# map_dbl(seq(1, 2, 0.1), function(n_mult){
#   map_dbl(1:100, function(h){
#     pc_comp_df %>%
#       filter(depth=="Deep") %>%
#       group_by(sla_class) %>%
#       summarize(mean=mean(frac_in_metab), sd=sd(frac_in_metab), n=n()) %>%
#       mutate(sim_vals=map(1:3, function(i){rnorm(.$n[i]*n_mult, .$mean[i], .$sd[i])})) %>%
#       unnest_longer(sim_vals) %>%
#       aov(formula=sim_vals~sla_class) %>%
#       broom::tidy() %>%
#       filter(term=="sla_class") %>%
#       pull(p.value)
#   }) %>% mean()
# })

*Supplemental figure: Fraction of particulate carbon represented by known metabolites. Boxplots set behind points are colored according to sea level anomaly (SLA) with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone).*

This correlation between PC and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. Known metabolites contributed between approximately 1.5% and 5% of the carbon biomass to the total PC load, but this variance was not clearly due to either depth or SLA. The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.3-0.4 percentage points lower at depth, p-value=0.018) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). However, this lack of difference observed may have been due to the small sample sizes (18 anticyclonic samples, 9 cyclonic, and 6 neither). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for the 175 meter samples if 1.5x as many samples had been collected with mean and variances remaining similar.

@@

*Supplemental figure: Linear regressions of rank-order peak area against SLA showing strong enrichment at the cyclone DCM and the anticyclone 175 meter samples.*

### Compositional shifts in metabolome as well as bulk biomass

@@

At the DCM, twenty-two compounds changed significantly in relative peak area across the eddy dipole. Seven of these were known metabolites, of which trigonelline and homarine were most enriched in the cyclone while arsenobetaine was the only metabolite with a larger peak area fraction in the anticyclone. Homarine showed a very large and highly significant shift, representing about 2.5% of the total peak area in the anticyclone but 7-8% in the cyclone center. Trigonelline, an isomer structurally very similar but biologically entirely distinct, had a surprisingly strong correlation with homarine (Pearson's $r=.855$) with peak areas consistently around one third of homarine's but large enough to have the second-largest shift in relative peak area of the significantly different compounds. Similarly, arsenobetaine represented about 0.1% of the peak area in the cyclone and 0.5-0.6% of the total peak area in the anticyclone. The most significantly different compounds at the DCM in each direction, however, were both unknowns. The lowest p-value ($6.4*10^{-6}$ after FDR correction) in the DCM data was a mass feature with *m/z* of 173.09211, a retention time also around 8.6 minutes, and enrichment in the anticyclone with a putative chemical formula of [M+H] = C$_7$H$_{13}$N$_2$O$_3$, possibly glyclproline or prolylglycine. The next-lowest (p = $2.5*10^{-4}$ after FDR correction) was enriched in the cyclone and had an *m/z* of 275.07119 and a retention time of 11.4 minutes with a putative formula of C$_{18}$H$_{11}$O$_3$.

@@

### Eddy center DCM samples show differences between depths and eddy polarities

@@

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 4). We found that a majority of the metabolites (Clusters 1 and 2 in Figure 4, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 4). Cluster #2 notably had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster #1 also showed a neat DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (r=.851) that was not present in the cyclone (r=.013).

@@

### Falkor data reveals different community and response to SLA

@@

When analyzed separately as distinct depths, the surface samples had higher R$^2$ values (R$^2$=.26) and a lower p-value (0.013), while the DCM samples were no longer likely to be distinct between the two eddies (p-value = 0.171) (Supplemental figure XX).

*Supplemental figure(?): Plots of the metabolome in the Falkor data, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. SLA trends are visible in the surface samples, with dark blue circles consistently discriminating from the dark red circles. The bottom row of plots show the direction and magnitude of this effect by plotting the grand mean of the normalized metabolite peak areas in the stations taken across the eddy dipole.*

@@

>>>>

---
title: "MESOSCOPE manuscript"
author: "William Kumler"
date: "`r Sys.Date()`"
output: 
  word_document:
    reference_docx: template.docx
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, eval = FALSE)
library(tidyverse)

@@

# Abstract

# Introduction

Until recently, the ocean's subtropical gyres were considered homogenous and static [@Karl1999]. In the past 25 years, however, regular sampling and advances in measurement technology have revealed its variability on time scales from the diel to the seasonal [@Karl2017; @Karl2021]. A major source of this variability comes in the form of mesoscale eddies in which water is entrained into circular surface currents tens to hundreds of kilometers in diameter [@Karl2017; @McGillicuddy2016]. Prior research has established the importance of these mesoscale features in introducing nutrients into the euphotic zone and altering the community composition [@Sweeney2003; @Barone2019]. In particular, the deep chlorophyll maximum (DCM) is known to correspond to the top of the nitricline and differs dramatically between cyclonic and anticyclonic eddies in depth and intensity. @Barone2022 reported a shift in the microbial DCM community with eukaryotic phytoplankton enriched and prokaryotic phytoplankton depleted in cyclonic eddies with the converse true for anticyclonic ones, while @Hawco2021 and @Dugenne2023 showed that some of this shift was due to alterations in dissolved iron chemistry and nitrogen fixation, especially at the surface.

These recent works have highlighted the effect of eddies on both bulk oceanographic parameters such as particulate carbon and nitrogen and the composition of the microbial community, but we still lack direct links between the genomic and transcriptomic assays occurring regularly around the world and the particulate organic matter produced by those organisms [@Kharbush2020]. This missing link is bridged by metabolomics, a nascent field studying the structure and function of the cell on a molecular level. Of particular interest within the metabolome are those small, polar molecules that represent the fundamental unit of cell biology and function as building blocks for macromolecules [@Moran2022]. These small molecules also have roles of their own within the cell, acting as osmolytes, antioxidants, nutrient storage compounds, and more [@Geraldes2021; @Mojzes2020; @Kharbush2020]. Despite these critical functions, they are difficult to study in the marine environment due to their low concentrations and similarity to the abundant salt ions, resulting in challenges in detection and quantification [@Boysen2018; @Moran2022; @Sacks2022].

Here, we use spatially resolved measurements of metabolites in particles across an eddy dipole in the North Pacific Subtropical Gyre (NPSG) to categorize unidentified molecular features based on their co occurrence with a set of simultaneously measured physical, chemical, and biological parameters. The gyre in particular represents an intriguing opportunity for metabolomic research because it has been shown to contain the highest proportion of chemical features with unknown structure and function and have highly variable responses to changes in latitude, depth, and across diel cycles [@Heal2021; @Boysen2021; @Kumler2023]. This study connects biogeochemical measurements to a molecular-scale perspective of particulate organic matter, linking the chemical currencies that fuel metabolism and interactions among marine microbes to the mesoscale environment in which they’re found.

# Results
## Metabolome changes significantly across the eddy dipole at DCM and 175m

We found that the metabolome differed between the cyclone and anticyclone samples, with multivariate statistics detecting the largest differences in the DCM and 175 meter samples while the surface was more homogenous (Figure 1). At the DCM, approximately 16% of the variation in the metabolome could be explained by SLA alone according to the PERMANOVA (p = 0.003) and SLA was strongly correlated with the first principal component (r = .615, % variance explained by PC1 = 27.2), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy were highly distinct and likely drove much of the variance explained. In the deep samples, 12% of the variance was explained by SLA (p = 0.008) according to the PERMANOVA and SLA was highly correlated with the second PC of the dataset (r = .697, % variance explained by PC2 = 15.4), though the first PC did not seem to have any visible pattern with metadata and appeared to capture largely variation between biological triplicates (Supplemental figure 1). At the surface, SLA trends are less evident with a larger p-value (0.087) and a lower R$^2$ (0.09) (Figure 1).

@@

@@

```{r PC vs metab abund anovas and perm test, eval=FALSE}
pc_comp_df <- targ_peaks %>%
  filter(str_detect(filename, "180821_Smp")) %>%
  left_join(clean_stans %>% select(compound_name, formula, polarity)) %>%
  mutate(n_C=as.numeric(str_extract(formula, "(?<=C)\\d+"))) %>%
  mutate(nM_C=n_C*nM) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  filter(samp_type=="Smp") %>%
  group_by(filename) %>%
  summarize(total_nM=sum(nM_C), PC_um=unique(PC_um)) %>%
  mutate(frac_in_metab=total_nM/1000/PC_um) %>%
  left_join(filled_file_metadata, by = join_by(filename)) %>%
  mutate(depth=str_extract(depth, "DCM|Deep|Surface")) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  mutate(station=factor(station, levels=c(as.character(4:14))))

# Significance testing between depths and SLA classes for % PC in metab
pc_comp_df %>% aov(formula=frac_in_metab~depth) %>% TukeyHSD()
pc_comp_df %>% aov(formula=frac_in_metab~sla_class) %>% TukeyHSD()

# Significance testing for differences in % in metabs as a function of SLA by depth
pc_comp_df %>%
  nest(data=-depth) %>%
  mutate(aov_out=map(data, ~broom::tidy(aov(.x$frac_in_metab~.x$sla_class)))) %>%
  mutate(kw_out=map(data, ~broom::tidy(kruskal.test(.x$frac_in_metab~.x$sla_class)))) %>%
  pivot_longer(c(aov_out, kw_out)) %>%
  unnest(value) %>%
  drop_na(p.value) %>%
  select(depth, name, p.value) %>%
  group_by(name) %>%
  mutate(p.value=p.adjust(p.value, method="fdr")) %>%
  ungroup() %>%
  arrange(name, depth)
pc_comp_df %>%
  ggplot(aes(x=sla_class, y=frac_in_metab)) +
  geom_boxplot() +
  facet_wrap(~depth, nrow=1)

# Permutation testing for effect size test on Deep SLA differences
set.seed(123)
map_dbl(seq(1, 2, 0.1), function(n_mult){
  map_dbl(1:100, function(h){
    pc_comp_df %>%
      filter(depth=="Deep") %>%
      group_by(sla_class) %>%
      summarize(mean=mean(frac_in_metab), sd=sd(frac_in_metab), n=n()) %>%
      mutate(sim_vals=map(1:3, function(i){rnorm(.$n[i]*n_mult, .$mean[i], .$sd[i])})) %>%
      unnest_longer(sim_vals) %>%
      aov(formula=sim_vals~sla_class) %>%
      broom::tidy() %>%
      filter(term=="sla_class") %>%
      pull(p.value)
  }) %>% mean()
})
```

This correlation between PC and overall metabolite abundance implied that the fraction of PC represented by known metabolites was likely to be fixed across the eddy transect and this expectation was validated by followup analyses. Known metabolites contributed between approximately 1.5% and 5% of the carbon biomass to the total PC load, but this variance was not clearly due to either depth or SLA (Supplemental figure 2). The only difference we observed was a slightly reduced metabolite contribution in the 175 meter samples (0.3-0.4 percentage points lower at depth, p-value=0.018) while no difference was detected between the surface and DCM samples or between the cyclone, anticyclone, and outside the eddy samples (all p-values > 0.1). However, this lack of difference observed may have been due to the small sample sizes (18 anticyclonic samples, 9 cyclonic, and 6 neither). A followup permutational sensitivity analysis indicated that a difference in means of 0.4 percentage points would have been obtained at the α < 0.05 level for the 175 meter samples if 1.5x as many samples had been collected with mean and variances remaining similar.

@@

## Compositional shifts in metabolome as well as bulk biomass

@@

At the DCM, twenty-two compounds changed significantly in relative peak area across the eddy dipole. Seven of these were known metabolites, of which trigonelline and homarine were most enriched in the cyclone while arsenobetaine was the only metabolite with a larger peak area fraction in the anticyclone. Homarine showed a very large and highly significant shift, representing about 2.5% of the total peak area in the anticyclone but 7-8% in the cyclone center. Trigonelline, an isomer structurally very similar but biologically entirely distinct, had a surprisingly strong correlation with homarine (Pearson's $r=.855$) with peak areas consistently around one third of homarine's but large enough to have the second-largest shift in relative peak area of the significantly different compounds. Similarly, arsenobetaine represented about 0.1% of the peak area in the cyclone and 0.5-0.6% of the total peak area in the anticyclone. The most significantly different compounds at the DCM in each direction, however, were both unknowns. The lowest p-value ($6.4*10^{-6}$ after FDR correction) in the DCM data was a mass feature with *m/z* of 173.09211, a retention time also around 8.6 minutes, and enrichment in the anticyclone with a putative chemical formula of [M+H] = C$_7$H$_{13}$N$_2$O$_3$, possibly glycylproline or prolylglycine. The next-lowest (p = $2.5*10^{-4}$ after FDR correction) was enriched in the cyclone and had an *m/z* of 275.07119 and a retention time of 11.4 minutes with a putative formula of C$_{18}$H$_{11}$O$_3$.

@@

## Eddy center DCM samples show differences between depths and eddy polarities

@@

To characterize the observed differences in multivariate space we used k-means clustering as an unsupervised way to identify dominant trends within the data (Figure 4). We found that a majority of the metabolites (Clusters 1 and 2 in Figure 4, 63% of the total) fell into clusters with larger peak areas in the cyclone, as expected, while also detecting 32 metabolites that clustered such that the mean metabolite was enriched in the anticyclone (Cluster 4, Figure 4). Cluster 2 notably had a distinct decrease in relative peak area with depth, while the other clusters had much less clear depth trends. Cluster 1 also showed a neat DCM maximum for the samples from the anticyclone, correlating well with flow cytometry counts of picoeukaryotes (r=.851) that was not present in the cyclone (r=.013).

@@

## Falkor data reveals different community and response to SLA

@@

When analyzed separately as distinct depths, the surface samples had higher R$^2$ values (R$^2$=.26) and a lower p-value (0.013), while the DCM samples were no longer likely to be distinct between the two eddies (p-value = 0.171) (Supplemental figure 3).

@@

# Discussion

# Methods
# Conclusions
# References

<div id="refs"></div>

# Supplement
## Supplemental figure 1

*Supplemental figure 1: Results of a principal component analysis performed separately on each depth of the eddy transect metabolome. A) Regressions of SLA against various principal components, broken down by depth. % variance explained by each PC is noted in the upper right corner of the plot. Colors have been added corresponding to the SLA value for clarity but are redundant with the x-axis values. Lines of best fit and the associated standard error have been added behind the points in black and translucent grey, respectively. Strong associations exist between PC1 and SLA in the DCM samples and PC2 and SLA in the deep samples. B) Principal component 1 values for the 175 meter samples at each station showing large variance between triplicates rather than any external factor.*

## Supplemental figure 2

*Supplemental figure 2: Fraction of particulate carbon represented by known metabolites. Boxplots set behind points are colored according to sea level anomaly (SLA) with red corresponding to a positive SLA (anticyclone) and blue corresponding to negative SLA (cyclone).*

## Supplemental figure 3

*Supplemental figure 3: Plots of the metabolome in the Falkor data, broken down by depth. The top row of plots are non-metric multidimensional scaling (NMDS) plots, in which points correspond to individual samples and have been colored by their corrected sea level height anomaly. SLA trends are visible in the surface samples, with dark blue circles consistently discriminating from the dark red circles. The bottom row of plots show the direction and magnitude of this effect by plotting the grand mean of the normalized metabolite peak areas in the stations taken across the eddy dipole.*

## Unnumbered figures, location/inclusion TBD

*Supplemental figure? Linear regressions of rank-order peak area against SLA showing strong enrichment at the cyclone DCM and the anticyclone 175 meter samples.*
